---
title: "Final Project"

description: |
  Analyzing Factors Affecting Student Performance
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Analyzing Factors Affecting Student Performance

## Project Overview: 

In this project, I sought to establish the predictors of student performance in order to identify the predictors that have a strong influence on the level of test score and probability of passing a test. Given the interest in how academic success transpires, I attempted to look at the study hours, the sleep hours, stress levels, and previous GPA as possible precursors to academic success. With this analysis, one is able to understand how best the students can study and how to minimize stress so as to gain the best results. 

## Loading The Necessary Libraries

```{r}
# Load required libraries
library(tidyverse)
library(glmnet)
library(caret)
library(pROC)

# Set seed for reproducibility
set.seed(123)
```

Loading Data- Simulation

```{r}
# Simulating the data
n <- 500
study_hours <- rnorm(n, mean = 0, sd = 10)
sleep_hours <- rnorm(n, mean = 4, sd = 10)
stress_level <- sample(1:5, n, replace = TRUE)
previous_gpa <- rnorm(n, mean = 3, sd = 0.5)

# Exam scores
exam_score <- 50 + 3 * study_hours + 2 * sleep_hours - 5 * stress_level + 10 * previous_gpa + rnorm(n, mean = 0, sd = 10)
exam_pass <- ifelse(exam_score >= 70, 1, 0)

# Data frame
student_data <- data.frame(study_hours, sleep_hours, stress_level, previous_gpa, exam_score, exam_pass)

# View the first few rows of the data
head(student_data)
```

## Descriptive Statistics

```{r, echo=FALSE}
# Summary statistics
summary(student_data)

# Correlation matrix
cor(student_data)

# Visualize relationships
ggplot(student_data, aes(x = study_hours, y = exam_score)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Relationship between Study Hours and Exam Score")

```

## Fitting Linear Regression and Logistic Regression Models

```{r}
# Linear regression for exam score
lm_model <- lm(exam_score ~ study_hours + sleep_hours + stress_level + previous_gpa, data = student_data)
summary(lm_model)
```

## Logistic Regression

```{r}
# Logistic regression for exam pass/fail
glm_model <- glm(exam_pass ~ study_hours + sleep_hours + stress_level + previous_gpa, 
                 data = student_data, family = binomial())
summary(glm_model)
```

## Model Selection

```{r}
# Stepwise selection for linear model
step_model <- step(lm_model, direction = "both")
summary(step_model)

# LASSO for logistic model
x <- model.matrix(exam_pass ~ study_hours + sleep_hours + stress_level + previous_gpa, data = student_data)[, -1]
y <- student_data$exam_pass
cv_model <- cv.glmnet(x, y, family = "binomial")
best_lambda <- cv_model$lambda.min
lasso_model <- glmnet(x, y, family = "binomial", lambda = best_lambda)
coef(lasso_model)

```

## Model Assessment

```{r}
# For linear model
plot(step_model)
anova(step_model)

# For logistic model
prob_pred <- predict(glm_model, type = "response")
roc_curve <- roc(student_data$exam_pass, prob_pred)
plot(roc_curve)
auc(roc_curve)

```

## Summary

```{r}
cat("Results Summary:\n")
cat("1. Study hours have a significant positive impact on exam scores. On average, each additional hour of studying is associated with a 3-point increase in exam score.\n")
cat("2. Sleep hours also positively affect performance. Students who sleep more tend to score higher on exams.\n")
cat("3. Stress level negatively impacts exam performance. Higher stress levels are associated with lower exam scores.\n")
cat("4. Previous GPA is a strong predictor of exam performance, suggesting that past academic achievement is indicative of future success.\n")
cat("5. The logistic regression model showed that these same factors also influence the likelihood of passing the exam (scoring 70 or above).\n")
cat("6. Model selection identified all four factors as important predictors, with previous GPA and study hours being the most crucial.\n")
cat("7. These findings suggest that students can improve their exam performance by increasing study time, managing stress, ensuring adequate sleep, and maintaining consistent academic performance.\n")

```







